Dependent Object Types

### 2012-09-11 doc/candidacy/slides.tex
tribe rev
nuobj rev

\begin{frame}{Tribe Calculus: Virtual Classes}
\begin{itemize}
\item {\it families} of classes inherited together
\item classes are lexically nested inside other classes: when a class
  is inherited, its nested inner classes are inherited along with
  their methods and fields
\item two form of inheritance: subclassing and further binding
\item {\it family polymorphism}: code written for one family
  also works for extensions of that family
\item in Tribe:\begin{itemize}
\item path types can depend simultaneously on both classes and objects
\item paths can use an {\tt out} field to move from an object to the
  object which surrounds it; enables ubiquitous access to an object's
  family without the need to drag around family arguments
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Tribe: Example}
\begin{minted}{java}
class Graph {
  class Node {
    Edge connect(Node other) {
      return new Edge(this, other);
    }
  }
  class Edge {
    Node from, to;
    Edge(Node f, Node t) { from = f; to = t; }
  }
}

@@

\begin{frame}[fragile]{Tribe: Example (continued)}
\begin{minted}{java}
final ColouredGraph cg1, cg2;
cg1.Node cn1, cn3;
cg2.Node cn2;
cn1.connect(cn3); // Type Correct
cn2.connect(cn3); // Type Error!!!

@@

%maybe one slide on implementation?

\begin{frame}{Tribe: Discussion}
\begin{itemize}
\item no virtual classes in Scala; its virtual types can emulate some (but not all) of the benefits
\item because of {\tt out} field, Tribe has limited cross-family inheritance %example?
\item soundness? substitution lemma seems wrong when {\tt null} is
  substituted for a variable in an expression whose type contains
  that variable %details?
\item aims to have decidable type-checking, but still not proved so
\end{itemize}
\end{frame}

@@

\begin{frame}{${\nu}Obj$: Syntax}
\begin{itemize}
\item terms in ${\nu}Obj$ denote objects or classes; consist of:
\begin{description}[mixin compositions]
\item[variables] $x$
\item[selections] $t.l$
\item[object creations] \mbox{$\newin{x \larrowc t}{u}$}
\item[class templates] $\cls{x: S \sep \tup d}$
\item[mixin compositions] $t_1 \withsub{S} t_2$\\(if omitted, $S = {S_1 \with (S_2 \with
  \obj{x \sep D_1 \uplus D_2})}$)
\end{description}
\item a value is a variable or class template
\item a path is a variable followed by a possibly empty sequence of selections
\item types in ${\nu}Obj$ consist of:
\begin{description}[mixin compositions]
\item[singleton types] $p.\type$
\item[type selections] $T\Tsel L$
\item[record types] $\obj{x \sep \tup D}$ % explain difference between record and class types?
\item[compound types] $T \with U$
\item[class types] $\cls{x: S \sep \tup D}$
\end{description}
\end{itemize}
\end{frame}

@@

% can you make this slide less cluttered?
\begin{frame}[fragile]{${\nu}Obj$: Example (encoding of monomorphic functions)}
A $\lambda$-abstraction $\lambda ( x : T )\; t$ is represented as a
class with an abstract member \code!arg! for the function argument:
\begin{lstlisting}
[x: {arg: T}| fun = [res = $t'$]]
\end{lstlisting}
$t'$ corresponds to term $t$ in which all occurrences of $x$ are
replaced by \code!$x$.arg!. We cannot directly access \code!arg! on
the right-hand side of \code!fun!: this wouldn't be ``contractive'',
because \code!arg! might not be initialized before \code!fun!. Hence,
we pack $t'$ in another class to ensure contractiveness. Now, an
application $g(e)$ gets decomposed into three subsequent steps:
\begin{lstlisting}
$\nu\,$g$_{app}$  <- g & [arg = e];
$\nu\,$g$_{eval}$ <- g$_{app}$.fun;
g$_{eval}$.res
\end{lstlisting}
First we instantiate function \code!g! with a concrete argument
yielding a thunk \code!g$_{app}$!. Then we evaluate this thunk by
creating an instance \code!g$_{eval}$! of it. Finally we extract the
result by querying field \code!res! of \code!g$_{eval}$!.
\end{frame}

\begin{frame}[fragile]{${\nu}Obj$: Discussion}
\begin{itemize}
\item conflate the concepts of compound types (which inherit the
  members of several parent types) and mixin composition (which build
  classes from other classes and traits)
\item mixin composition is not commutative, unlike classical
  intersection types
\item in Scala, least upper bounds and greatest lower bounds do not always exist, e.g: %can you translate this to nuObj?
\end{itemize}
\begin{minted}{scala}
  trait A { type T <: A }
  trait B { type T <: B }
  // glb is approximated by an infinite sequence
  A with B { type T <: A with B { type T <: A with B {
    type T <: ...
  }}}
\end{minted}
\end{frame}

>>>>

\begin{frame}[fragile]{Tribe: Example (Virtual Classes)}
\begin{minted}{java}
class Graph {
  class Node {
    Edge connect(Node other) {
      return new Edge(this, other);
    }
  }
  class Edge {
    Node from, to;
    Edge(Node f, Node t) { from = f; to = t; }
  }
}

@@

\begin{frame}[fragile]{Tribe: Example (Object Family and Family Polymorphism)}
\begin{minted}{java}
final ColouredGraph cg1, cg2;
cg1.Node cn1, cn3;
cg2.Node cn2;
cn1.connect(cn3); // Type Correct
cn2.connect(cn3); // Type Error!!!

@@

\begin{frame}{Tribe Calculus: Recap}
\begin{itemize}
\item {\it families} of classes inherited together
\item classes are lexically nested inside other classes: when a class
  is inherited, its nested inner classes are inherited along with
  their methods and fields
\item two form of inheritance: subclassing and further binding
\item {\it family polymorphism}: code written for one family
  also works for extensions of that family
\item in Tribe:\begin{itemize}
\item path types can depend simultaneously on both classes and objects
\item paths can use an {\tt out} field to move from an object to the
  object which surrounds it; enables ubiquitous access to an object's
  family without the need to drag around family arguments
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Tribe: Discussion}
\begin{itemize}
\item no virtual classes in Scala; its virtual types can emulate some (but not all) of the benefits
\item because of {\tt out} field, Tribe has limited cross-family inheritance %example?
\begin{minted}[fontsize=\footnotesize]{java}
class A {
  class D { ... }
  class B {
    class C {
      this.out.out.D foo;
    }
  }
}
class E {
  class F extends A.B.C { ... }
}
\end{minted}
\item soundness? substitution lemma seems wrong when {\tt null} is
  substituted for a variable in an expression whose type contains
  that variable %details?
\item aims to have decidable type-checking, but still not proved so
\end{itemize}
\end{frame}

@@

\begin{frame}{${\nu}Obj$: Syntax}
\begin{itemize}
\item terms in ${\nu}Obj$ denote objects or classes; consist of:
\begin{description}[mixin compositions]
\item[variables] $x$
\item[selections] $t.l$
\item[object creations] \mbox{$\newin{x \larrowc t}{u}$}
\item[class templates] $\cls{x: S \sep \tup d}$
\item[mixin compositions] $t_1 \withsub{S} t_2$\\(if omitted, $S = {S_1 \with (S_2 \with
  \obj{x \sep D_1 \uplus D_2})}$)
\end{description}
\item a value is a variable or class template
\item a path is a variable followed by a possibly empty sequence of selections
\item types in ${\nu}Obj$ consist of:
\begin{description}[mixin compositions]
\item[singleton types] $p.\type$
\item[type selections] $T\Tsel L$
\item[record types] $\obj{x \sep \tup D}$ % explain difference between record and class types?
\item[class types] $\cls{x: S \sep \tup D}$
\item[compound types] $T \with U$
\end{description}
\end{itemize}
\end{frame}

@@

\begin{frame}[fragile]{${\nu}Obj$: Example (encoding of monomorphic functions)}
\begin{itemize}
\item Encoding for a $\lambda$-abstraction $\lambda ( x : T )\; t$:
\begin{lstlisting}[basicstyle=\large]
[x: {arg: T}| fun = [res = $t'$]]
\end{lstlisting}
($t'$ = $t$ with \code{$x$.arg} substituted for $x$)

\item Encoding for an application $g(e)$:
\begin{lstlisting}[basicstyle=\large]
$\nu\,$g$_{app}$  <- g & [arg = e];
$\nu\,$g$_{eval}$ <- g$_{app}$.fun;
g$_{eval}$.res
\end{lstlisting}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{${\nu}Obj$: Discussion}
\begin{itemize}
\item conflate the concepts of compound types (which inherit the
  members of several parent types) and mixin composition (which build
  classes from other classes and traits)
\item mixin composition is not commutative, unlike classical
  intersection types
\item in Scala, least upper bounds and greatest lower bounds do not always exist, e.g: %can you translate this to nuObj?
\end{itemize}
\begin{minted}{scala}
  trait A { type T <: A }
  trait B { type T <: B }
  // glb is an infinite sequence
  A with B { type T <: A with B { type T <: A with B {
    type T <: ...
  }}}
\end{minted}
\end{frame}
